Body size patterns

### 2019-03-22 MS_Analysis_3spp_4rivers_Patterns_from 2spp.Rmd
added graphs

modCSRY <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = riverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortSpeciesRiverOrdered) +
               #s(year, bs = 'cr', by = speciesOrdered), 
               s(year, bs = 'cr', k = kIn, by = speciesFactor),
               data = dataIn, 
               method = "REML")
summary(modCSRY)
plot(modCSRY, pages=10, scale = 0)

predGridCSRY <- expand.grid(cohortSpeciesRiverOrdered = unique(modCSRY$model$cohortSpeciesRiverOrdered),
                            year = unique(modCSRY$model$year),
                            ydayCumul = seq(250,1000, by = 30)) %>%
                    mutate(cohortOrdered = as.numeric(substr(cohortSpeciesRiverOrdered,1,4)),
                           speciesOrdered = substr(cohortSpeciesRiverOrdered,6,8),
                           riverOrdered = substring(cohortSpeciesRiverOrdered,10),
                           cohortRiverOrdered = paste(cohortOrdered, riverOrdered, sep = '_'),
                           speciesCohortOrdered = paste(speciesOrdered, cohortOrdered, sep = '_'),
                           speciesRiverOrdered = paste(speciesOrdered, riverOrdered, sep = '_'),
                           cohortCentered = cohortOrdered - cohortOffset,
                           cohort = cohortOrdered,
                           species = speciesOrdered,
                           river = riverOrdered)

predCSRY <- predict(modCSRY, newdata = predGridCSRY)#, type = 'lpmatrix')

predGrid2CSRY <- bind_cols(predGridCSRY,as.data.frame(predCSRY))

ggplot(predGrid2CSRY, aes(ydayCumul, predCSRY, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(50,300) +
  facet_grid(river ~ cohortOrdered) +
  ggtitle('modCSRY')

# trying a couple of models
modCSRC <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = riverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortSpeciesRiverOrdered) +
               #s(year, bs = 'cr', by = speciesOrdered), 
               s(cohort, bs = 'cr', k = kIn, by = speciesFactor),
               data = dataIn, 
               method = "REML")
summary(modCSRC)
plot(modCSRC, pages=8, scale = 0)

modC_S_RC <- bam(observedLength ~ cohort + species + river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = riverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortSpeciesRiverOrdered) +
               s(cohort, bs = 'cr', k = kIn, by = speciesFactor),
               data = dataIn, 
               method = "REML")
summary(modC_S_RC)
plot(modCSRC, pages=8, scale = 0)

@@

save(mod,modCR,modCSR,modSC,modSR,modCSRY,modC_S_R, file = './data/out/mods.RData')
AIC(modC_S_RC,modC_S_R,modCSR,modSR,modSC,modCR,mod)  %>% rownames_to_column() %>% arrange(AIC)

>>>>

# modCSRY <- bam(observedLength ~ cohort * species * river + 
#                s(ydayCumul, bs = 'cr') +
#                s(ydayCumul, bs = 'cr', k = kIn, by = cohortOrdered) +
#                s(ydayCumul, bs = 'cr', k = kIn, by = speciesOrdered) +
#                s(ydayCumul, bs = 'cr', k = kIn, by = riverOrdered) +
#                s(ydayCumul, bs = 'cr', k = kIn, by = cohortRiverOrdered) +
#                s(ydayCumul, bs = 'cr', k = kIn, by = speciesCohortOrdered) +
#                s(ydayCumul, bs = 'cr', k = kIn, by = speciesRiverOrdered) +
#                s(ydayCumul, bs = 'cr', k = kIn, by = cohortSpeciesRiverOrdered) +
#                #s(year, bs = 'cr', by = speciesOrdered), 
#                s(year, bs = 'cr', k = kIn, by = speciesFactor),
#                data = dataIn, 
#                method = "REML")
# summary(modCSRY)
# plot(modCSRY, pages=10, scale = 0)
# 
# predGridCSRY <- expand.grid(cohortSpeciesRiverOrdered = unique(modCSRY$model$cohortSpeciesRiverOrdered),
#                             year = unique(modCSRY$model$year),
#                             ydayCumul = seq(250,1000, by = 30)) %>%
#                     mutate(cohortOrdered = as.numeric(substr(cohortSpeciesRiverOrdered,1,4)),
#                            speciesOrdered = substr(cohortSpeciesRiverOrdered,6,8),
#                            riverOrdered = substring(cohortSpeciesRiverOrdered,10),
#                            cohortRiverOrdered = paste(cohortOrdered, riverOrdered, sep = '_'),
#                            speciesCohortOrdered = paste(speciesOrdered, cohortOrdered, sep = '_'),
#                            speciesRiverOrdered = paste(speciesOrdered, riverOrdered, sep = '_'),
#                            cohortCentered = cohortOrdered - cohortOffset,
#                            cohort = cohortOrdered,
#                            species = speciesOrdered,
#                            river = riverOrdered)
# 
# predCSRY <- predict(modCSRY, newdata = predGridCSRY)#, type = 'lpmatrix')
# 
# predGrid2CSRY <- bind_cols(predGridCSRY,as.data.frame(predCSRY))
# 
# ggplot(predGrid2CSRY, aes(ydayCumul, predCSRY, color = species)) +
#   geom_point() + 
#   geom_line() +
#   ylim(50,300) +
#   facet_grid(river ~ cohortOrdered) +
#   ggtitle('modCSRY')

# trying a couple of models
modCSR_C <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = riverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortSpeciesRiverOrdered) +
               s(cohort, bs = 'cr', k = kIn, by = speciesFactor),
               data = dataIn, 
               method = "REML")
summary(modCSR_C)
plot(modCSR_C, pages=10, scale = 0)

@@

save(mod,modCR,modCSR,modSC,modSR,modC_S_R,modCSR_C, file = './data/out/mods.RData')
AIC(modCSR_C,modC_S_R,modCSR,modSR,modSC,modCR,mod)  %>% rownames_to_column() %>% arrange(AIC)

### 2019-03-22 MS_Analysis_3spp_4rivers_Patterns_from 2spp_bySpp.Rmd
added graphs

```{r gams for body size - model selection}
speciesIn <- 'bnt'
depVarIn <- 'observedLength'
kIn <- 4 # # of knots for the speces and river smooths

@@

# number of knots from gam 
kValue <- 9 # from mod6 below, just for graphing

# by species (bkt, bnt), by cohort, by river gridded
ggplot(vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt', 'bnt', 'ats')), aes(ydayCumul,observedLength, group = species, color = (species))) +
  geom_point(size = 0.5, alpha = 0.1) +
 # geom_smooth(span = 0.1, size = 1.5, se = F) +
  geom_smooth(method='gam', alpha=0.1, formula=y~s(x, k = kValue), se = F) +
  facet_grid(riverOrdered ~ cohort) +
  theme_publication()

####################################
# gam models

@@

dataIn <- vB %>% filter(cohort %in% 2000:2013, species %in% speciesIn)
kIn <- 4
################################################################################

@@

```{r best models}

@@

modBNT <- bam(observedLength ~ cohort * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = riverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortRiverOrdered) +
               s(cohort, bs = 'cr', k = kIn), 
             data = vB %>% filter(cohort %in% 2000:2013, species %in% 'bkt'), method = "REML")

@@

ggplot(predGrid2Both, aes(ydayCumul, pred, group = cohort, color = cohort)) +
  geom_point() + 
  geom_line() +
  labs(y = 'Predicted body size (mm)',
       x = 'Days since age-0 Jan 1',
       color = 'Cohort') +
  scale_color_viridis_c(breaks = c(2000,2004,2008,2012), labels = c(2000,2004,2008,2012)) +
  theme_publication() +
  theme(legend.position = "right", 
        legend.direction = "vertical",
        legend.title = element_text(face="plain")) +
  guides(color = guide_colorbar(barheight = 10,
                                barwidth = 1.5,
                                ticks = FALSE,
                                nbin = 100)) +
#  guide_legend(title = 'Cohort') +
  facet_grid(speciesGG ~ riverGG)

>>>>

```{r gams for body size - data prep}
speciesIn <- 'bnt'
depVarIn <- 'observedLength'
kIn <- 4 # # of knots for the speces and river smooths

@@

@@

dataIn <- vB %>% filter(cohort %in% 2000:2013, species %in% speciesIn)
kIn <- 4
```

Run to here to start

```{r raw data graphs}
# number of knots from gam 
kValue <- 9 # from mod6 below, just for graphing

# by species (bkt, bnt), by cohort, by river gridded
ggplot(vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt', 'bnt', 'ats')), aes(ydayCumul,observedLength, group = species, color = (species))) +
  geom_point(size = 0.5, alpha = 0.1) +
 # geom_smooth(span = 0.1, size = 1.5, se = F) +
  geom_smooth(method='gam', alpha=0.1, formula=y~s(x, k = kValue), se = F) +
  facet_grid(riverOrdered ~ cohort) +
  theme_publication()
```

```{r gam models}
####################################
# gam models

################################################################################

@@

```{r best models for each species}

@@

modBNT <- bam(observedLength ~ cohort * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = riverOrdered) +
               s(ydayCumul, bs = 'cr', k = kIn, by = cohortRiverOrdered) +
               s(cohort, bs = 'cr', k = kIn), 
             data = vB %>% filter(cohort %in% 2000:2013, species %in% 'bnt'), method = "REML")

@@

save(modBKT,modBNT,predBKT,predBNT,predGrid2Both, file = paste0('./data/out/bestModsBKTBNT.RData'))

```{r graphs of best models}
# load('./data/out/bestModsBKTBNT.RData')

# cohorts by river/species
ggplot(predGrid2Both, aes(ydayCumul, pred, group = cohort, color = cohort)) +
  geom_point() + 
  geom_line() +
  labs(y = 'Predicted body size (mm)',
       x = 'Days since age-0 Jan 1',
       color = 'Cohort') +
  scale_color_viridis_c(breaks = c(2000,2004,2008,2012), labels = c(2000,2004,2008,2012)) +
  theme_publication() +
  theme(legend.position = "right", 
        legend.direction = "vertical",
        legend.title = element_text(face="plain")) +
  guides(color = guide_colorbar(barheight = 10,
                                barwidth = 1.5,
                                ticks = FALSE,
                                nbin = 100)) +
#  guide_legend(title = 'Cohort') +
  facet_grid(speciesGG ~ riverGG, scales = 'free_y')

# specific splines
library(gratia)
cohortBKT <- evaluate_smooth(modBKT, "s(cohort)") %>% mutate(species = 'bkt')
cohortBNT <- evaluate_smooth(modBNT, "s(cohort)") %>% mutate(species = 'bnt')
cohortBoth <- bind_rows(cohortBKT,cohortBNT) %>% mutate(upperCI = est + se * 1.96, lowerCI = est - se * 1.96)

cohortBoth$speciesGG <- factor(cohortBoth$species, levels = c('bkt','bnt','ats'), labels = c("Brook Trout", "Brown Trout", "Atlantic Salmon"), ordered = T)

ggplot(cohortBoth, aes(cohort, group = speciesGG)) + 
  geom_ribbon(aes( ymin = lowerCI, ymax = upperCI), fill  = 'grey70') +
  geom_line(aes(y = est, linetype = speciesGG), size = 1.5) +
  labs(y = 'Effect size',
       x = 'Cohort',
       linetype = 'Species') +
  theme_publication() +
  theme(legend.position = "right", 
        legend.direction = "vertical",
        legend.title = element_text(face="plain"))

# main spline over ydayCumul
plotBKT = plot.gam(modBKT, select = 1)
plotBNT = plot.gam(modBNT, select = 1)

plotBKTydayCumul <- data.frame(x = plotBKT[[1]]$x, y = plotBKT[[1]]$fit, se = plotBKT[[1]]$se) %>% mutate(species = 'bkt')
plotBNTydayCumul <- data.frame(x = plotBNT[[1]]$x, y = plotBNT[[1]]$fit, se = plotBNT[[1]]$se) %>% mutate(species = 'bnt')
plotBothydayCumul <- bind_rows(plotBKTydayCumul,plotBNTydayCumul) %>% mutate(upperCI = y + se * 1.96, lowerCI = y - se * 1.96)

plotBothydayCumul$speciesGG <- factor(plotBothydayCumul$species, levels = c('bkt','bnt','ats'), labels = c("Brook Trout", "Brown Trout", "Atlantic Salmon"), ordered = T)

ggplot(plotBothydayCumul, aes(x, group = speciesGG)) + 
  geom_ribbon(aes( ymin = lowerCI, ymax = upperCI, fill  = speciesGG), alpha = 0.5) +
  geom_line(aes(y = y, linetype = speciesGG), size = 1.5) +
  scale_fill_manual(values = c('grey30', 'grey70')) +
  labs(y = 'Predicted body size (mm)',
       x = 'Days since age-0 Jan 1',
       linetype = 'Species') +
  theme_publication() +
  theme(legend.position = "right", 
        legend.direction = "vertical",
        legend.title = element_text(face="plain"))

# main

ggplot(tmp2, aes(x,y)) + geom_line()

appraise(modBKT)
